---
title: "Property association"
output: flexdashboard::flex_dashboard
vertical_layout: scroll
date: '`r Sys.Date()`'
params:
  metadata: # vector of all metadata files paths
  taxa: # vector of all taxa values files paths
  join: # rows or columns, indicate the direction to join metadatas
  new_property: "" # list of vectors with the name of the new property, the property to be used to create the new one, the method to create the new one
  cat_properties: # vector of categorical properties 
  cont_properties: # vector of continuous properties 
  pca2prop: "" # list of 3 vectors, 2 of them with the properties to be compared in same scatter plot and the last with the palette name for the first property
  picture_3d: "" # list of vectors with the properties (prop), the angles (angle) in picture 3d and the 3 components (PCs)
  outlier_thr: 0.95 # indicate threshold for outlier removal
  values_not_considered: # vector of values not to be considered in wilcoxon/kruskal tests for cat feat
  taxa_selection: "all" # vector of taxa (column name) to be selected
  taxa_selection_approach: "join" # join, intersect, setdiff, indicate the way combination of different taxa to be selected
  labels_to_be_used: "" # vector with labels to be used for heatmap
  total_taxa: "../../data/all_bacteria_species.txt" # path where all taxa are listed
  rotations_path: "" # path to print out first 6 pca loadings ordered by importance
  pca_matrix_path: "" # path to print out pca matrix
  heatmap_ptab_path: "" # path to print out p values heatmap
  heatmap_qtab_path: "" # path to print out q values heatmap
  heatmap_nPCs: "" # number of PCs to be considered for the table output
  bubble_plot: "" # if bubble plot should be drawn
  n_feat: 1000 # number of microbes to be selected
  palette: "" # name of one of my palette in general_function
---

```{r setup, include = FALSE, dev = "CairoPNG"}
	knitr::opts_chunk$set(dev="CairoPNG")
```

```{r libraries, include = FALSE}
	library(DT)
	library(knitr)
	library(evaluate)
	library(scatterplot3d)
```

```{r functions, include = FALSE}
	source("../general_functions.R")
	source("functions.R")
```	

```{r data, include = FALSE}
	# Upload of
		# Metadata
			if(length(params$metadata)>1){
				full_metadata=join_metadata(metadata_paths=params$metadata, join_by=params$join, matching1="file_id", matching2="file_id")
			} else {
				full_metadata=read.csv(params$metadata, sep="\t", header=TRUE, check.names=FALSE, stringsAsFactors=FALSE)
			}
		# Taxa quantification
			if(length(params$taxa)>1){
				taxa_values=join_taxa_table(taxa_paths=params$taxa)
			} else {
				taxa_values=read.csv(params$taxa, sep="\t", header=TRUE, check.names=FALSE, stringsAsFactors=FALSE)
			}
	# Property modifications (must be column labels of full_metadata)
		cat_properties=params$cat_properties
		cont_properties=params$cont_properties
		if(all(cat_properties=="")){
			cat_properties=NA
		}
		if(all(cont_properties=="")){
			cont_properties=NA
		}
		n_properties=length(cat_properties[!is.na(cat_properties)])+length(cont_properties[!is.na(cont_properties)])
		not_considered=params$values_not_considered
		if(length(not_considered)==""){
			not_considered=NA
		}
		if(length(cat_properties)!=length(not_considered)){
			stop("Problem in length cat properties and values_not_considered")
		}
		names(not_considered)=cat_properties
	# Details for plot
		# 2 properties scatterplots
			if(params$pca2prop==""){
				pca2prop=NA
			} else {
				pca2prop=params$pca2prop
			}
		# 3D versions of scatterplots
			if(params$picture_3d==""){
				picture_3d=NA
			} else {
				picture_3d=params$picture_3d
			}
	# Set the palette
		if(params$palette!=""){
			if(all(!is.na(cat_properties)) & all(!is.na(cont_properties))){
				palette_col=my_palettes[params$palette]
				names(palette_col)=c(cat_properties, cont_properties)
			} else if(all(is.na(cat_properties)) & all(!is.na(cont_properties))){
				palette_col=my_palettes[params$palette]
				names(palette_col)=cont_properties
			} else if(all(!is.na(cat_properties)) & all(is.na(cont_properties))){
				palette_col=my_palettes[params$palette]
				names(palette_col)=cat_properties
			}
		} else {
			if(all(!is.na(cat_properties)) & all(!is.na(cont_properties))){
				palette_col=rep("default", length(cat_properties)+length(cont_properties))
				names(palette_col)=c(cat_properties, cont_properties)
			} else if(all(is.na(cat_properties)) & all(!is.na(cont_properties))){
				palette_col=rep("default", length(cont_properties))
				names(palette_col)=cont_properties
			} else if(all(!is.na(cat_properties)) & all(is.na(cont_properties))){
				palette_col=rep("default", length(cat_properties))
				names(palette_col)=cat_properties
			}
		}	
	# Total taxa
		all_taxa_tab=read.csv(params$total_taxa, sep="\t", header=TRUE, check.names=FALSE, stringsAsFactors=FALSE)		
		all_taxa=all_taxa_tab[,2]
		names(all_taxa)=as.character(all_taxa_tab[,1])
```

```{r select_samples, include = FALSE}
	# Match metadata and taxa_values
		metadata=match_metadata(metadata=full_metadata, taxa_values=taxa_values, colname_metadata="file_id")
```

```{r taxa_sel, include = FALSE}
	# Select a subset of taxa
	if(all(params$taxa_selection!="all")){
		subset_taxa_list=lapply(params$taxa_selection, function(x){
				subset_taxa_tab=read.csv(x, sep="\t", header=TRUE, check.names=FALSE, stringsAsFactors=FALSE)		
				subset_taxa=subset_taxa_tab[,2]
				names(subset_taxa)=as.character(subset_taxa_tab[,1])
				return(subset_taxa)
			}
		)
		if(all(params$taxa_selection_approach=="join")){
			subset_taxa=lapply(subset_taxa_list, names)
			subset_taxa=unique(unlist(subset_taxa))
		} else if(all(params$taxa_selection_approach=="intersect")){
			subset_taxa_tmp=lapply(subset_taxa_list, names)
			subset_taxa=list_intersection(subset_taxa_tmp)
		} else if(length(params$taxa_selection)==1 & params$taxa_selection_approach=="setdiff"){
			subset_taxa_tmp=lapply(subset_taxa_list, names)
			subset_taxa=setdiff(colnames(taxa_values), subset_taxa_tmp[[1]])
		} else if(all(params$taxa_selection_approach=="setdiff")){
			subset_taxa_tmp=lapply(subset_taxa_list, names)
			subset_taxa=list_setdiff(subset_taxa_tmp)
		} else {
			subset_taxa_tmp=names(subset_taxa_list[[1]])
			for(i in 1:length(params$taxa_selection_approach)){
				if(params$taxa_selection_approach[[i]]=="join"){
					tmp=unique(c(subset_taxa_tmp, names(subset_taxa_list[[i+1]])))
					subset_taxa_tmp=tmp
				} else if(params$taxa_selection_approach[[i]]=="intersect"){
					tmp=intersect(subset_taxa_tmp, names(subset_taxa_list[[i+1]]))
					subset_taxa_tmp=tmp
				} else if(params$taxa_selection_approach[[i]]=="setdiff"){
					tmp=setdiff(subset_taxa_tmp, names(subset_taxa_list[[i+1]]))
					subset_taxa_tmp=tmp
				} 
			}
			subset_taxa=subset_taxa_tmp
		}
		
		# select taxa 
		taxa_values=taxa_values[,subset_taxa]
	}
```

```{r new_metadata_property, include = FALSE}
	# Make new properties 
		if(all(params$new_property!="")){
			for(i in 1:length(params$new_property)){
				feat=create_new_property(metadata=metadata, old_feat=params$new_property[[i]][["old"]], met=params$new_property[[i]][["met"]])
				metadata=data.frame(metadata, feat)
				colnames(metadata)[ncol(metadata)]=params$new_property[[i]][["new_name"]]
			}
		}
```

```{r PCA, echo = FALSE}
	# PCA
		pca_data=PCA(taxa_values=taxa_values, out_removal=params$outlier_thr, num_features=params$n_feat)

	# PCA rotations
		pca_rotations=lapply(paste("PC", 1:6, sep=""), function(x){
				val=pca_data[["rotation"]][,x]
				val=abs(val)
				thr_val=sort(val, decreasing=TRUE)
				tax=names(thr_val)
				nam_tax=all_taxa[tax]
				nam_tax[is.na(nam_tax)]=tax[is.na(nam_tax)]
				names(nam_tax)=NULL
				res=data.frame(tax, nam_tax)
				colnames(res)=c(x, paste("name_", x, sep=''))
				rownames(res)=NULL
				return(res)
			}
		)
		rotations_tab=do.call(cbind, pca_rotations)
		if(params$rotations_path!=""){
			write.table(rotations_tab, file=params$rotations_path, quote=FALSE, sep="\t", row.names=FALSE, col.names=TRUE)
		}

	# PCA matrix
		tab=data.frame(rownames(pca_data$x), pca_data$x)
		colnames(tab)[1]="file_id"
		if(params$pca_matrix_path!=""){
			write.table(tab, file=params$pca_matrix_path, quote=FALSE, sep="\t", row.names=FALSE, col.names=TRUE)
		}

	# PCA plots
		pca_plots=PCA_plots(pca_data=pca_data, metadata=metadata, cat_properties=cat_properties, 
													cont_properties=cont_properties, PCs=1:6, val_removed=not_considered,
													pal=palette_col)
	# PCA plots 2 properties
		if(any(!is.na(pca2prop))){
			if(pca2prop[[3]]!="brew" & pca2prop[[3]]!="default"){
				palette_tmp=lapply(pca2prop[[3]], function(x){
						my_palettes[[pca2prop[[3]]]]
					}
				)
			} else {
				palette_tmp=pca2prop[[3]]
			}
			pca_plots_2prop=PCA_plots_2properties(pca_data=pca_data, metadata=metadata, PCs=1:6, 
																						prop1=pca2prop[[1]], prop2=pca2prop[[2]],
																						pal=palette_tmp)
		} else {
			pca_plots_2prop=NA
		}
```

```{r ass_cor_PCA_to_properties, echo = FALSE}
	# Test associations and correlations of properties to PCs
		PCA_ass_cor=test_dimreduction_to_features(dimred_data=pca_data$x, metadata=metadata, cat_features=cat_properties, 
																								cont_features=cont_properties, val_removed=not_considered)
	
	# Heatmap
		heatmap_p_PCA_plot=feat_ass_heatmap(PCA_ass_cor=PCA_ass_cor, PCs=1:6, stat_val="p")
		heatmap_q_PCA_plot=feat_ass_heatmap(PCA_ass_cor=PCA_ass_cor, PCs=1:6, stat_val="q")

	# write table of p values
		if(params$heatmap_ptab_path!=""){
			if(!all(is.na(cat_properties))){
				tab1=sapply(PCA_ass_cor[[1]], function(x){
						if(any(colnames(x)=="krusk_p")){
							return(x[,"krusk_p"])
						} else if(any(colnames(x)=="wilc_p")){
							return(x[,"wilc_p"])
						}
					}
				)
			} else {
				tab1=NA
			}
			if(!all(is.na(cont_properties))){
				tab2=sapply(PCA_ass_cor[[2]], function(x){
						x[,"spearm_p"]
					}
				)
			} else {
				tab2=NA
			}
			if(is.na(tab1)){
				tab=tab2
			} else if(is.na(tab2)){
				tab=tab1
			} else {
				tab=cbind(tab1, tab2)
			}
			if(params$heatmap_nPCs!=""){
				tab=tab[1:params$heatmap_nPCs,]
			}
			tab=data.frame(PCs=rownames(tab), tab)
			write.table(tab, file=params$heatmap_ptab_path, quote=FALSE, sep="\t", row.names=FALSE, col.names=TRUE)
		}
	# write table of q values (correcting all the PCs)
		if(params$heatmap_qtab_path!="" & params$heatmap_nPCs==""){
			if(!all(is.na(cat_properties))){
				tab1=sapply(PCA_ass_cor[[1]], function(x){
						if(any(colnames(x)=="krusk_q")){
							return(x[,"krusk_q"])
						} else if(any(colnames(x)=="wilc_q")){
							return(x[,"wilc_q"])
						}
					}
				)
			} else {
				tab1=NA
			}
			if(!all(is.na(cont_properties))){
				tab2=sapply(PCA_ass_cor[[2]], function(x){
						x[,"spearm_q"]
					}
				)
			} else {
				tab2=NA
			}
			if(is.na(tab1)){
				tab=tab2
			} else if(is.na(tab2)){
				tab=tab1
			} else {
				tab=cbind(tab1, tab2)
			}
			tab=data.frame(PCs=rownames(tab), tab)
			write.table(tab, file=params$heatmap_qtab_path, quote=FALSE, sep="\t", row.names=FALSE, col.names=TRUE)
		}
	# write table of q values (correcting a defined number of PCs)
		if(params$heatmap_qtab_path!="" & params$heatmap_nPCs!=""){
			if(!all(is.na(cat_properties))){
				tab1=sapply(PCA_ass_cor[[1]], function(x){
						if(any(colnames(x)=="krusk_p")){
							return(x[,"krusk_p"])
						} else if(any(colnames(x)=="wilc_p")){
							return(x[,"wilc_p"])
						}
					}
				)
			} else {
				tab1=NA
			}
			if(!all(is.na(cont_properties))){
				tab2=sapply(PCA_ass_cor[[2]], function(x){
						x[,"spearm_p"]
					}
				)
			} else {
				tab2=NA
			}
			if(is.na(tab1)){
				tab=tab2
			} else if(is.na(tab2)){
				tab=tab1
			} else {
				tab=cbind(tab1, tab2)
			}
			tab=tab[1:params$heatmap_nPCs,]
			row_names=rownames(tab)
			col_names=colnames(tab)
			tab=round(p.adjust(tab, method="fdr"), 2)
			tab=data.frame(row_names, matrix(tab, nrow=params$heatmap_nPCs))
			colnames(tab)=c("PCs", col_names)
			write.table(tab, file=params$heatmap_qtab_path, quote=FALSE, sep="\t", row.names=FALSE, col.names=TRUE)
		}	
```

```{r pca_3D, echo = FALSE}
	if(any(!is.na(picture_3d))){
		three_dim_pict_path=c()
		for(i in 1:length(picture_3d)){
			# Remove from metadata the samples considered outliers
				p=sapply(rownames(pca_data$x), function(x){
						which(metadata[,"file_id"]==x)
					}
				)
				metadata_3d=metadata[p,]

			# Percentage of variability explained
				eigs=pca_data$sdev^2
				perc=eigs/sum(eigs)*100

			for(y in picture_3d[[i]][["feat"]]){
				# define colors
					lev=levels(metadata_3d[,y])
					if(is.null(lev)){
						lev=unique(metadata_3d[,y])
					}
					if(all(palette_col[[picture_3d[[i]][["feat"]]]]=="default")){
						shades=gg_color_hue(length(lev))
					} else {
						shades=palette_col[[picture_3d[[i]][["feat"]]]]
					}
					names(shades)=lev
					col=sapply(rownames(pca_data$x), function(t){
				      proj=metadata_3d[metadata_3d[,"file_id"]==t,y]
				      return(shades[proj])
				    }
				  )
				  col=adjustcolor(col, alpha.f = 0.5)
				# Legend
					num_observations=table(metadata_3d[,y])
					pos=which(num_observations!=0 & num_observations!="logical")
					num_observations=num_observations[pos]
					new_labels=paste(names(num_observations), "(", num_observations, ")", sep='')
					tmp=data.frame(pca_data$x[,1:2], metadata_3d[,y])
					colnames(tmp)=c("x_val", "y_val", "color_group")
					col2=shades
					names(col2)=NULL
					PCA_tmp=ggplot(tmp)+
										geom_point(size=2, aes(x=color_group, y="PC1",
														col=factor(color_group, labels=new_labels)))+
										labs(color=y)+ 
										scale_color_manual(values=col2)+
										theme(legend.title=element_text(size=20), legend.text=element_text(size=20))
					leg=get_legend(PCA_tmp)
					leg_plot=as_ggplot(leg)
					# save plot
				 	ggsave(filename=paste(params$picture_3d[[i]]$path, "_legend_PCA3D_", y, ".pdf", sep=""),
								  dpi=500, plot=leg_plot
								)
				# PCA 3D
					# invert the order
						tab=pca_data$x[,picture_3d[[i]]$PCs]
			 			tab=tab[nrow(tab):1,]
			 			col=col[length(col):1]
			 		# create a temporary file
				  	pdf(paste(params$picture_3d[[i]]$path, "_PCA3D_", y, "_PC", picture_3d[[i]]$PCs[[1]], 
													picture_3d[[i]]$PCs[[2]], picture_3d[[i]]$PCs[[3]], ".pdf", sep=''))
						  dev.control(displaylist="enable")
						  scatter3d_plot=scatterplot3d(x=pca_data$x[,picture_3d[[i]]$PCs[[1]]], 
														y=pca_data$x[,picture_3d[[i]]$PCs[[2]]], 
														z=pca_data$x[,picture_3d[[i]]$PCs[[3]]], 
														color="white", angle=picture_3d[[i]]$angle, box=FALSE, grid=TRUE, 
														xlab=paste("PC", picture_3d[[i]]$PCs[[1]], sep=''), 
														ylab=paste("PC", picture_3d[[i]]$PCs[[2]], sep=''),
														zlab=paste("PC", picture_3d[[i]]$PCs[[3]], sep=''),
														cex.axis=1.5,
														cex.lab=1.7,
														pch=16
														)
							addgrids3d(pca_data$x[,picture_3d[[i]]$PCs], grid=c("xy", "xz"), angle=picture_3d[[i]]$angle)
					 		scatter3d_plot$points3d(tab, pch=19, col=col)
					  	PCA3D_plot=recordPlot()
				  	dev.off()
				# Add the file path
			  three_dim_pict_path=c(three_dim_pict_path, 
			  											paste(params$picture_3d[[i]]$path, "_legend_PCA3D_", y, ".pdf", sep=""),
			  											paste(params$picture_3d[[i]]$path, "_PCA3D_", y, "_PC", picture_3d[[i]]$PCs[[1]], picture_3d[[i]]$PCs[[2]], picture_3d[[i]]$PCs[[3]], ".pdf", sep='')
			  										)
			}
		}
	}
```

```{r bubble_plot, echo = FALSE}
	# bubbleplot
	if(params$bubble_plot!=""){
		cont_ass_list=lapply(1:length(PCA_ass_cor[[2]]), function(x){
				res=PCA_ass_cor[[2]][[x]][1:6, c("spearman_r", "spearman_p")]
				res=data.frame(res, p_values=res[,2], PCs=paste("PC", 1:6, sep=""), feat=rep(names(PCA_ass_cor[[2]])[x], 6), lab=signif(res[,"spearman_r"], 2))
				return(res)
			}
		)
		cont_ass_tab=do.call(rbind, cont_ass_list)
		cont_ass_tab[cont_ass_tab[,"p_values"]>0.05,"p_values"]=NA
		bubble_plot=ggplot(cont_ass_tab, aes(x=PCs, y=feat, 
           									colour=p_values, size=spearman_r)) +
								  geom_point() +
								  geom_text(aes(label=lab), colour="black", size=3) +
								  scale_size_continuous(range=c(5, 15), name="Spearman R") + 
								  theme_light() +
									scale_color_gradient2(low = "red",
																			  high = "blue", name="p values") +
								  labs(x = NULL, y = NULL) 
	} else {
		bubble_plot=NA
	}
```

```{r num_microbes, echo = FALSE}
	tot_microbes=length(which(apply(taxa_values, 2, function(x){!all(x==0)})))
```


General {data-orientation=rows}
===================================== 
	
Total number of taxa detected: `r tot_microbes` \
Total number of taxa used in PCA: `r length(pca_data$center)` \
Number of sample outliers: `r length(setdiff(rownames(taxa_values), rownames(pca_data$x)))`

```{r, results='asis', fig.show = 'asis', out.width="25%"}
	pca_plots[['scree']]
	pca_plots[['cum_scree']]
	cat(knitr::knit_print(kable(rotations_tab[1:50,grep("name_", colnames(rotations_tab))], format.args=list(digits=20, scientific=TRUE), "simple")))
```

PCA: scatter plots {data-orientation=rows}
===================================== 

Row {.tabset .tabset-fade}
-------------------------------------

```{r, results='asis', fig.show = 'asis', out.width="25%"}
	out = NULL
	if(!is.na(cat_properties)){
		for(i in 1:length(cat_properties)){
			cat("\n", sep="")
			cat("### ", names(pca_plots[["cat_plot"]])[i], "\n", sep="")
			if(!is.na(pca_plots[["cat_plot"]][[i]][['legend']])){
				evaluate(print(pca_plots[["cat_plot"]][[i]][['legend']]))
				evaluate(print(pca_plots[["cat_plot"]][[i]][['scatters']]))
			}
			cat("\n", sep="")
		}
	}
	if(!is.na(cont_properties)){
		for(i in 1:length(cont_properties)){
			cat("\n", sep="")
			cat("### ", names(pca_plots[["cont_plot"]])[i], "\n", sep="")
			if(!is.na(pca_plots[["cont_plot"]][[i]][['legend']])){
				evaluate(print(pca_plots[["cont_plot"]][[i]][['legend']]))
				evaluate(print(pca_plots[["cont_plot"]][[i]][['scatters']]))
			}
			cat("\n", sep="")
		}
	}
```

PCA: two properties scatter plots {data-orientation=rows}
===================================== 

Row {.tabset .tabset-fade}
-------------------------------------

```{r, results='asis', fig.show = 'asis', out.width="25%"}
	out = NULL
	if(!is.na(pca_plots_2prop)){
		for(i in 1:length(pca2prop[[1]])){
			cat("\n", sep="")
			cat("### ", pca2prop[[1]][i], " ", pca2prop[[2]][i], "\n", sep="")
			evaluate(print(pca_plots_2prop[[i]][['legend']]))
			evaluate(print(pca_plots_2prop[[i]][['scatter']]))
			cat("\n", sep="")
		}
	}
```

PCA: 3D scatterplots {data-orientation=rows}
===================================== 

```{r, results='asis'}
	out = NULL
	if(!is.na(picture_3d)){
		for(i in 1:length(three_dim_pict_path)){
			cat("\n", sep="")
			cat(paste("![](", three_dim_pict_path[[i]], "){width=500px height=500px}", sep=""))
		}
	}
```

PCs {data-orientation=rows}
===================================== 

Row {.tabset .tabset-fade}
-------------------------------------

```{r, results='asis', fig.show = 'asis', out.width="25%"}
	out = NULL
	if(!is.na(cat_properties)){
		for(i in 1:length(cat_properties)){
			cat("\n", sep="")
			cat("### ", names(pca_plots[["cat_plot"]])[i], "\n", sep="")
			if(!is.na(pca_plots[["cat_plot"]][[i]][['by_PCs']])){
				evaluate(print(pca_plots[["cat_plot"]][[i]][['by_PCs']]))
			}
			cat("\n", sep="")
		}
	}
	if(!is.na(cont_properties)){
		for(i in 1:length(cont_properties)){
			cat("\n", sep="")
			cat("### ", names(pca_plots[["cont_plot"]])[i], "\n", sep="")
			if(!is.na(pca_plots[["cont_plot"]][[i]][['by_PCs']])){
				evaluate(print(pca_plots[["cont_plot"]][[i]][['by_PCs']]))
			}
			cat("\n", sep="")
		}
	}
```

Heatmap {data-orientation=rows}
===================================== 

Row {.tabset .tabset-fade}
-------------------------------------

```{r, results='asis', fig.show = 'asis', fig.dim = c(7, n_properties/2.53), out.width="50%"}
	heatmap_p_PCA_plot
	heatmap_q_PCA_plot
```

PCA: bubble_plot {data-orientation=rows}
===================================== 

Row {.tabset .tabset-fade}
-------------------------------------

```{r, results='asis', fig.show = 'asis', out.width="50%"}
	if(!is.na(bubble_plot)){
		bubble_plot
	}
```
